Data exploration through dot-driven development\footnote{This work was supported by Google Digital News Initiative.}

### 2017-01-01 pivot/paper.tex
Relational algebra

@@

If $\Gamma\vdash e:[f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]$ using row types and
$\Gamma\vdash e:C$ then $\ident{fields}(C) = \{ f_1\mapsto\tau_1 \}$
\begin{equation*}
\end{equation*}

@@

\subsection{Relational algebra}
\begin{equation*}
\begin{array}{rclll}
  R & = & \Pi_{f_1, \ldots, f_n} (R) &&\textnormal{get columns}\\
   & \lsep & \sigma_\varphi (R) &&\textnormal{filter by predicate}\\
   & \lsep & \tau_{f_1, \ldots, f_n}(R) &&\textnormal{sort by fields}\\
   & \lsep & \Phi_{f, \rho_1, \ldots, \rho_n} (R) &&\textnormal{group by $f$ and aggregate}\\
   \\
  \rho & = & \ident{count}  \\
   & \lsep & \ident{sum}~f    \\       
   & \lsep & \ident{dist}~f\\
   & \lsep & \ident{conc}~f\\
\end{array}
\end{equation*}
\subsection{Foo calculus}
\begin{equation*}
\begin{array}{rcl}
 \tau &=& \ident{num} \lsep \ident{string} \lsep \ident{series}\langl \tau_1, \tau_2 \rangl \lsep \ident{Query}\\
 l &=& \kvd{type}~C(\overline{x:\tau}) = \overline{m} \\[0.0em]
 m &=& \kvd{member}~N:\tau= e
\end{array}
\end{equation*}

\begin{equation*}
\inference[(foo)]
  {L_1; \Gamma \vdash e : C; L_2 \qquad ((\kvd{type}~C(\overline{x:\tau}) = ..\;\kvd{member}~N_i : \tau_i = e_i\;..), L) \in L_2}
  {L_1; \Gamma \vdash e.N_i:\tau_i; L_2 \cup L}
\end{equation*}

>>>>

\theoremstyle{plain}
\theoremstyle{definition}
\newtheorem{rmrk}{Remark}

@@

In our approach, the information about available fields is used by a type provider to provide
types with appropriate members. This is hidden from the host language, which only sees class types
with members. Provided class definitions consist of a constructor and members: 
%
\begin{equation*}
\begin{array}{rcl}
 l &=& \kvd{type}~C(x:\tau) = \overline{m} \\
 m &=& \kvd{member}~N:\tau= e
\end{array}
\end{equation*}

\noindent
During type-checking, the type system keeps track of a list of provided class definitions $L$. 
Checking member access is then just a matter of finding the corresponding class definition and
finding the member type:
%
\begin{equation*}
\inference[(\emph{member})~]
  {L; \Gamma \vdash e : C & \kvd{type}~C(x:\tau) = ..\;\kvd{member}~N_i : \tau_i = e_i\;.. \in L}
  {L; \Gamma \vdash e.N_i:\tau_i}
\end{equation*}

\noindent
The rule, adapted from [X], does not capture laziness of type providers that is 
important for the pivot type provider (where the number of provided classes is potentially 
infinite). We discuss this aspect in Section~\ref{sec:foo}.

Using type providers and nominal type system seemingly hides knowledge about fields available
in the data set. However, for types constructed by the pivot type provider, we can define a 
mapping \ident{fields} that returns the fields available in the data set represented by the 
class. The type provider encodes the logic expressed in Section~\ref{sec:columns-row} in the
following sense:

\begin{rmrk}[Encoding of fancy types]
\label{thm:encoding-fancy}
If $\Gamma\vdash e:[f_1\!:\!\tau_1, \ldots, f_n\!:\!\tau_n]$ using a type system defined in 
Figure~\ref{fig:fancy-types} and $\Gamma\vdash e:C$ using nominal typing and $C$ is a type 
provided by the pivot type provider then $\ident{fields}(C) = \{ f_1\mapsto\tau_1, \ldots, f_n\mapsto\tau_n \}$.
\end{rmrk}

\noindent
In the following two sections, we focus on formalizing the pivot type provider and the nominally 
typed host language. We define the \ident{fields} predicate in Section~\ref{sec:proofs} and use it 
to prove properties of the pivot type provider.

We do not fully develop the type system based on fancy types sketched in Section~\ref{sec:columns-row}.
However, the remark illustrates one interesting aspect of our work -- the type provider 
mechanism makes it possible to express safety guarantees that would normally require row types and 
type state in a simple nominally typed language. In a similar way, type providers have been used
to encode session types [X], suggesting that this is a generally useful approach.

@@

\subsection{Relational algebra}
\begin{equation*}
\begin{array}{rclll}
  R & = & \Pi_{f_1, \ldots, f_n} (R) &&\textnormal{get columns}\\
   & \lsep & \sigma_\varphi (R) &&\textnormal{filter by predicate}\\
   & \lsep & \tau_{f_1, \ldots, f_n}(R) &&\textnormal{sort by fields}\\
   & \lsep & \Phi_{f, \rho_1, \ldots, \rho_n} (R) &&\textnormal{group by $f$ and aggregate}\\
   \\
  \rho & = & \ident{count}  \\
   & \lsep & \ident{sum}~f    \\       
   & \lsep & \ident{dist}~f\\
   & \lsep & \ident{conc}~f\\
\\
\end{array} 
\end{equation*}

\newcommand{\vect}[1]{\langl #1 \rangl}

$v=\vect{\{f_1\mapsto v_{1,1}, \ldots, f_n\mapsto v_{1,n}\},\ldots,\{f_1\mapsto v_{m,1}, \ldots, f_n\mapsto v_{m,n}\}}$

\begin{figure}
\begin{equation*}
\begin{array}{l}
\Pi_{f_{p(1)}, \ldots, f_{p(m)}} \{f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} \leadsto \\
  \quad \{f_{p(1)}\mapsto \vect{v_{p(1),1},\ldots,v_{p(1),r}}, \ldots, f_{p(m)}\mapsto \vect{v_{p(m),1},\ldots,v_{p(m),r}}\}  
\\
\\
\sigma_\varphi \{f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} \leadsto\\
  \quad \{f_1\mapsto \vect{\ldots, v_{1,j}, \ldots}, \ldots, f_n\mapsto \vect{\ldots, v_{n,j},\ldots}\}
  \qquad (\forall j.\;\varphi\, \{f_1\mapsto v_{1,j}, \ldots, f_n\mapsto v_{n,j}\})
\\
\\
\tau_{f_{p(1)}, \ldots, f_{p(m)}} \{f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} \leadsto\\
  \quad \{f_1\mapsto \vect{v_{1,q(1)},\ldots,v_{1,q(r)}}, \ldots, f_n\mapsto \vect{v_{n,q(1)},\ldots,v_{n,q(r)}}\} 
  \quad \textnormal{where $q$ permutation}\\
  \qquad \textnormal{such that}~i\leq j\implies(v_{p(1),q(i)}, \ldots, v_{p(m),q(i)}) \leq (v_{p(1),q(j)}, \ldots, v_{p(m),q(j)})
  \\
  \\
\Phi_{f_i, \rho_1, \ldots, \rho_m} \{f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} \leadsto\\
  \quad \{\ident{nm}(\rho_1)\mapsto a_1, \ldots, \ident{nm}(\rho_m)\mapsto a_m \} 
  \quad \textnormal{where}\\
  \qquad \{g_1,\ldots,g_s\}=\{\{l~|~k\in 1\ldots r, ~v_{i,l}=v_{i,k}\}, ~l\in 1\ldots r\} \\
  ~~\begin{array}{ll}    
    \qquad a_i = \vect{|g_1|, \ldots, |g_s|}  & \textnormal{when}~\rho_i = \ident{count} \\
    \qquad a_i = \vect{\Sigma_{k\in g_1}v_{j,k}, \ldots, \Sigma_{k\in g_s}v_{j,k}} & \textnormal{when}~\rho_i = \ident{sum}~f_j \\
    \qquad a_i = \vect{\Pi_{k\in g_1}v_{j,k}, \ldots, \Sigma_{k\in g_s}v_{j,k}}   & \textnormal{when}~\rho_i = \ident{conc}~f_j \\
    \qquad a_i = \vect{|\{v_{j,k}, k\in g_1\}|, \ldots, |\{v_{j,k}, k\in g_s\}|}  & \textnormal{when}~\rho_i = \ident{dist}~f_j
  \end{array}
\end{array}
\end{equation*}

\caption{foo}
\label{fig:zzz}
\end{figure}

%\noindent
\begin{equation*}
\Pi_{f_{p(1)}, \ldots, f_{p(n)}} \{f_1\mapsto \overline{v_1}, \ldots, f_n\mapsto \overline{v_m}\}
  \leadsto \{f_{p(1)}\mapsto \overline{v_{p(1)}}, \ldots, f_{p(n)}\mapsto \overline{v_{p(n)}}\}
\end{equation*}

\subsection{Foo calculus}
\begin{equation*}
\begin{array}{rcl}
 \tau &=& C \lsep \ident{num} \lsep \ident{string} \lsep \ident{series}\langl \tau_1, \tau_2 \rangl \lsep \ident{Query}\\
 l &=& \kvd{type}~C(x:\tau) = \overline{m} \\[0.0em]
 m &=& \kvd{member}~N:\tau= e
\end{array}
\end{equation*}

\begin{equation*}
\begin{array}{rcl}
 v &\narrow{=}& d \lsep \ident{None} \lsep \ident{Some}(v) \lsep \kvd{new}~C(\overline{v}) \lsep v_1::v_2 \\[0.0em]
 e &\narrow{=}& d \lsep op \lsep e_1~e_2 \lsep \lambda x.e \lsep e.N \lsep \kvd{new}~C(\overline{e}) \\
   &\narrow{|}& \ident{None} \lsep\kvd{match}~e~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\
   &\narrow{|}& \ident{Some}(e) \lsep e_1 = e_2 \lsep {\kvd{if}~e_1~\kvd{then}~e_2~\kvd{else}~e_3} \lsep \kvd{nil} \\
   &\narrow{|}& e_1 :: e_2 \lsep \kvd{match}~e~\kvd{with}~x_1::x_2 \rightarrow e_1 \,|\, \kvd{nil} \rightarrow e_2
\end{array}
\end{equation*}

\begin{equation*}
\inference[(foo)]
  {L_1; \Gamma \vdash e : C; L_2 \qquad ((\kvd{type}~C(x:\tau) = ..\;\kvd{member}~N_i : \tau_i = e_i\;..), L) \in L_2}
  {L_1; \Gamma \vdash e.N_i:\tau_i; L_2 \cup L}
\end{equation*}
